Review Paper Draft

### 2020-08-15 main.tex
Sections 2.2 and 2.3 edits

An additional consideration in genetic studies that we will briefly mention is that of genetic background. \anna{make the following sentence better: correlated snps are the issue - talk about ld pruning} Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citet{vilhjalmsson2012nature} \pb{Use citet, not citep here} \anna{double check all citations for whether they should be (1) citet, citep or (2) better at the end of the sentence} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they? \textbf{LD pruning} or block-wise analysis, gene-level test, \textbf{haplotype block}...group correlated variants}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

@@

\anna{switch first and second paragraphs... the current first paragraph doesn't describe GC}

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. In order to reduce confounding these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in Figure \ref{fig:ps_env_block}. These methods do not directly address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs. Whether or not they are effective depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects. \anna{maybe say something about how 'if it's a good proxy - great! they'll work', but 'if it's not a good proxy, also fine! because then it's not a confounder! ...SE's potentially a problem like in GLS... but this is different with lasso. (maybe not worth getting into here)}

\anna{make a better slashed arrow}

@@

One of the first methods to address thie problem of confounding due to population structure is the genomic control (GC) method. GC is a univariate testing procedure which modifies a test statistic by a deflation factor. This deflation factor attempts to quantify and remove the effect of population structure on the distribution of the appropriate test statistic, and is estimated using loci independent of the phenotype of interest and the loci being tested for association \citep{devlin1999genomic, bacanu2000power, wang2009testing}. However, this deflation factor has been reported to be sensitive to the nature and quantity of the loci used to estimate it \citep{hellwege2017population, marchini2004effects}. Additionally, GC does not correct effect size estimates, so resulting coefficient estimates will be unreliable after GC is applied, even if their corresponding test statistics and p-values have been appropriately adjusted \citep{hellwege2017population}.

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their more fine-grained, subtle relatedness.

>>>>

An additional source of confounding in genetic studies that we will briefly mention is that of confounding due to genetic background. This occurs in univariate association methods when a non-causal SNP is tested for association with a trait of interest, and that SNP is correlated with a causal SNP. As \citet{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. LD pruning is commonly employed to reduce confounding due to genetic background in univariate association testing. However, we will focus on methods which address this via genomewide multivariate analysis.

@@

One of the first methods to address the problem of confounding due to population structure is the genomic control (GC) method. GC is a univariate testing procedure which modifies a test statistic by a deflation factor. This deflation factor attempts to quantify and remove the effect of population structure on the distribution of the appropriate test statistic, and is estimated using loci independent of the phenotype of interest and the loci being tested for association \citep{devlin1999genomic, bacanu2000power, wang2009testing}. However, this deflation factor has been reported to be sensitive to the nature and quantity of the loci used to estimate it \citep{hellwege2017population, marchini2004effects}. Additionally, GC does not correct effect size estimates, so resulting coefficient estimates will be unreliable after GC is applied, even if their corresponding test statistics and p-values have been appropriately adjusted \citep{hellwege2017population}.

New statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. Many of these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in Figure \ref{fig:ps_env_block}. These methods do not directly address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs.

\anna{make better slashed arrows / DAGs}

@@

Whether or not such methods are effective at reducing confounding effects depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects. In the presence of unobserved heterogeneous environmental effects which are not well-approximated by inferred population structure, these effects will not confound SNP-phenotype associations. This is due to the absence of a causal pathway between population stratification and environment, as depicted in Figure \ref{fig:ps_env_block2}. In such a scenario, failing to account for unobserved environmental effects will result in greater phenotypic variability \anna{and as a result, greater SE in beta-hats... if we're talking about gls}, however, it will not bias SNP estimates \anna{cite something from causal inference here}. We will return to this idea and its mathematical justification in section \ref{sec:methods}.

\begin{figure}[H]
\centering
\begin{tikzpicture}[%
>=latex',
circ/.style={draw, shape=circle, node distance=1cm, line width=1pt}]%Define the arrow type and style for circled nodes
\draw[dashed, red] (0,1) node[left] (P) {Population Stratification} --node {X} (2,1) node[right] (E) {Environment};
\draw[->] (0,0) node[left] (X) {Observed SNPs} -- (2,0) node[right] (Y) {Phenotype}; %Line between X and Y
\draw[->] (P) -- (X) ; 
\draw[->] (E)   -- (Y) ; 
% \filldraw[color=red!60, fill=none, very thick] (0,1) ellipse (4.5 and 0.35);
\end{tikzpicture}
\caption{The causal pathway between the environment and observed SNPs may be blocked if genetic information does not serve as an accurate proxy for it.}
\label{fig:ps_env_block2}
\end{figure}

\anna{need transition to bring it back to the first DAG for now...} Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their more fine-grained, subtle relatedness.
